Reduced growth due to belowground sink limitation is not fully explained by reduced photosynthesis

### 2017-02-27 manuscript/manuscript_pve_editor.Rmd
manuscript resubmitted
started revision of editor comments #2

Sink limitation is known to reduce plant growth, but it is not known how plant carbon (C) balance is affected, limiting our ability to predict growth under sink-limited conditions. We manipulated soil volume to impose sink limitation of growth in *Eucalyptus tereticornis* seedlings. Seedlings were grown in the field in containers of different sizes and planted flush to the soil alongside freely-rooted (Free) seedlings. Container volume negatively affected aboveground growth throughout the experiment, and light saturated rates of leaf photosynthesis were consistently lower in seedlings in containers (-26 %) compared to Free seedlings. Significant reductions in photosynthetic capacity in containerized seedlings was related to both reduced leaf nitrogen content and starch accumulation, indicating direct effects of sink limitation on photosynthetic down regulation. After 120 days, harvested biomass of Free seedlings was on average 84 % higher than seedlings in containers, but biomass distribution in leaves, stems and roots was not different. However, the reduction in net leaf photosynthesis over the growth period was insufficient to explain the reduction in growth, so that we also observed an apparent reduction in whole-plant C use efficiency (CUE) with container size. Our results show that sink limitation affects plant growth through feedbacks to both photosynthesis and CUE. Mass balance approaches to predicting plant growth under sink-limited conditions need to incorporate both of these feedbacks.

@@

\
*P* was predicted by using a coupled photosynthesis - stomatal conductance model [@farquhar1980biochemical;@medlyn2011reconciling] with the 'plantecophys' package [@duursma2015] in R with the mean photosynthetic parameters (*J*~max~, *V*~cmax~, *g*~1~ and *R*~dark~) for each treatment and meteorological data from an onsite weather station. Values of *J*~max~ and *V*~cmax~ are mean estimates from AC~i~ curves over two measurement periods (explained above). The *g*~1~ parameter was generated by fitting the optimal stomatal conductance model from [@medlyn2011reconciling] to observed *g*~s~ values. *R*~dark~ was was estimated using Eq. (2) from @atkin2005response, which was parameterized with a mean reference empirical measurement for each treatment, a Q~10~ value of 1.86, a reference air temperature for each treatment and air temperatures at each time step.

\
(2)
$$R~dark~ = R_{ref}Q_{10}~^{(T-T_{ref})/10}$$
Methods of the coupled leaf gas exchange model are also described in Duursma *et al.* [-@duursma2014peaked]. Combined with the meteorological parameters; PPFD, air temperature, and relative humidity, at 15 min intervals, net leaf photosynthesis rates ($\mu$mol CO~2~ m^-2^ s^-1^) were then predicted for each soil volume treatment. *P*~day~ was calculated by converting predicted rates to mass C gain over 15 min time steps (g m^-2^) and then summed for 24 h. This resulted in 120 unique values of *P* for each soil volume treatment, one value for each day of the experiment. Thus, each daily time step for model runs included a value of *P*~day~ that represented both treatment specific photosynthetic parameters and meteorological constraints across the duration of the experiment.

@@

\
We used modelled C gain  to test the hypothesis that the effects of belowground sink limitation on rates of leaf photosynthesis were sufficient to accurately predict overall seedling biomass production after 120 days. Cumulative net leaf C gain for each treatment was equal to the sum of each value of *P* over 120 days and final seedling C was assumed to equal half of the final dry mass for observed seedlings. We then compared seedling C use efficiency (CUE) across treatments, as the proportion of modelled cumulative net leaf C gain allocated to observed seedling biomass.

@@

##Growth and morphology
Plant height, diameter and leaf area diverged between container volumes soon after start of the experiment (Figure `r allomfig`a-c). First to diverge was seedling leaf area, which significantly differed among soil volumes (*P* < 0.029) from the 5^th^ week of the experiment. Significant differences were then observed in both height (8^th^ week) and then diameter (9^th^ week) among soil volumes (*P* < 0.045 & 0.035, respectively). The large reductions in height gain and total leaf area in smaller compared to larger containers continued throughout the experiment. In this field study, colder temperatures and reductions in total PPFD per day (Figure `r airvarsfig`) likely slowed the growth of seedlings in the final weeks of the experiment. Seedlings maintained diameter growth throughout the experiment, although growth was marginal in smaller soil volumes in the final month. Final seedling height significantly increased with increasing soil volume (*P* < 0.001). Increases in both final stem diameter and cumulative leaf area (both *P* < 0.001) were found with increasing soil volume. We did not find a continuous effect of container size; rather, plant size was similar among smaller containers, and differences were driven mainly by the largest container and the Free seedling treatments.

@@

\
Both *J*~max~ and *V*~cmax~ were significantly higher in Free seedlings (30 % and 26 %, respectively) than container-grown seedlings with little variation between container volume treatments (Table `r pvetab3`). Overall, the *J*~max~ / *V*~cmax~ ratio was conserved across all treatments (1.6&pm;0.02). Leaf dark respiration rates were not significantly different across soil volumes (Table `r pvetab3`). The *g*~1~ parameter, generated for each seedling from the Medlyn *et al.* [-@medlyn2011reconciling] optimal stomatal conductance model, was lowest in the Free seedling treatment and was marginally different across soil volume treatments (Table `r pvetab3`).

@@

##Whole-plant C balance
Modelled cumulative net leaf C gain over 120 days varied three-fold across containers sizes (Table `r pvetab1`). The Free seedling control had five-fold greater modelled total C gain than the average of container treatments (213.6 - 42.7&pm;8.2 g C). For seedlings in containers, an average of 67 &pm;0.01 % of modelled C gain was not allocated to observed biomass, compared to 59.2 % for the Free seedling control. Consequently, the estimated CUE of containerized seedlings was lower than the Free seedling control (Figure `r cuefig`). As a result, the observed reductions in leaf *A*~n~ with decreasing soil volume when integrated across the 120 day experiment were not large enough to explain the reduction in observed seedling biomass across the container size treatments.

@@

\
With high rates of *g*~s~, non-limiting leaf water potential ($\Psi$~pd~), and consistent leaf $\delta$^13^C across soil volume treatments, there was little evidence that water stress caused the reduction in *A*~sat~. This finding is consistent with other container size studies without drought treatments.  For example, reduced *A*~max~ in cotton seedlings grown at elevated CO~2~ was attributed to sink-limited feedback inhibition from inadequate rooting volume, not decreased *g*~s~ [@thomas1991root]. Similarly, severe reductions in leaf photosynthesis in coffee plants were not attributed to impacts of container size on leaf water potentials or *g*~s~ [@ronchi2006growth]. It is thus likely that reductions in *A*~at~ of well-watered seedlings observed in our study of *E. tereticornis* seedlings are not the result of limiting soil moisture availability.

@@

\
Focusing only on empirical results in this study points to *A*~n~ driving reductions in seedling growth. Modelling cumulative net seedling C gain, however, suggests that other components of the C balance, beyond *A*~n~, are required to explain the observed seedling biomass response to sink manipulation. This finding is noteworthy, as partial accounting of the different mechanisms involved in plant C balance can lead to erroneous conclusions [@valentini2000respiration]. For example, the fraction of photosynthate used in respiration is known to vary depending on species and local environmental conditions [@Lambers2008], yet is often considered a static parameter in process based growth models. Thus, we agree with conclusions from Valentine and M&#228;kel&#228; [-@valentine2005bridging] that the problem with predicting tree growth is a problem in forecasting the assimilation and allocation of C and other constituents. Future empirical and modelling studies should focus on how feedbacks from sink activity affect both rates of *A*~n~ and the fate of C allocated at the whole plant level to growth, respiration, exudation and C storage. However, the degree to which these mechanisms regulate the C available for growth will undoubtedly shift across different species and environmental conditions.

>>>>

Sink limitation is known to reduce plant growth, but it is not known how plant carbon (C) balance is affected, limiting our ability to predict growth under sink-limited conditions. We manipulated soil volume to impose sink limitation of growth in *Eucalyptus tereticornis* seedlings. Seedlings were grown in the field in containers of different sizes and planted flush to the soil alongside freely-rooted (Free) seedlings. Container volume negatively affected aboveground growth throughout the experiment, and light saturated rates of leaf photosynthesis were consistently lower in seedlings in containers (-26 %) compared to Free seedlings. Significant reductions in photosynthetic capacity in containerized seedlings was related to both reduced leaf nitrogen content and starch accumulation, indicating direct effects of sink limitation on photosynthetic down regulation. After 120 days, harvested biomass of Free seedlings was on average 84 % higher than seedlings in containers, but biomass distribution in leaves, stems and roots was not different. However, the reduction in net leaf photosynthesis over the growth period was insufficient to explain the reduction in growth, so that we also observed an apparent reduction in whole-plant C use efficiency (CUE) between Free seedlings and seedlings in containers. Our results show that sink limitation affects plant growth through feedbacks to both photosynthesis and CUE. Mass balance approaches to predicting plant growth under sink-limited conditions need to incorporate both of these feedbacks.

@@

\
*P*~day~  was predicted by using a coupled photosynthesis - stomatal conductance model [@farquhar1980biochemical;@medlyn2011reconciling] with the 'plantecophys' package [@duursma2015] in R with the mean photosynthetic parameters (*J*~max~, *V*~cmax~, *g*~1~ and *R*~dark~) for each treatment and meteorological data from an onsite weather station. Values of *J*~max~ and *V*~cmax~ are mean estimates from AC~i~ curves over two measurement periods (explained above). The *g*~1~ parameter was generated by fitting the optimal stomatal conductance model from [@medlyn2011reconciling] to observed *g*~s~ values. *R*~dark~ was was estimated using Eq. (2) from @atkin2005response, which was parameterized with a mean reference empirical measurement for each treatment, a Q~10~ value of 1.86, a reference air temperature for each treatment and air temperatures at each time step.

\
(2)
$$R~dark~ = R_{ref}Q_{10}~^{(T-T_{ref})/10}$$
Methods of the coupled leaf gas exchange model are also described in Duursma *et al.* [-@duursma2014peaked]. Combined with the meteorological parameters; PPFD, air temperature, and relative humidity, at 15 min intervals, net leaf photosynthesis rates ($\mu$mol CO~2~ m^-2^ s^-1^) were then predicted for each soil volume treatment. *P*~day~ was calculated by converting predicted rates to mass C gain over 15 min time steps (g m^-2^) and then summed for 24 h. This resulted in 120 unique values of *P*~day~  for each soil volume treatment, one value for each day of the experiment. Thus, each daily time step for model runs included a value of *P*~day~ that represented both treatment specific photosynthetic parameters and meteorological constraints across the duration of the experiment.

@@

\
We used modelled C gain  to test the hypothesis that the effects of belowground sink limitation on rates of leaf photosynthesis were sufficient to accurately predict overall seedling biomass production after 120 days. Cumulative net leaf C gain for each treatment was equal to the sum of each value of *P*~day~ over 120 days and final seedling C was assumed to equal half of the final dry mass for observed seedlings. We then compared seedling C use efficiency (CUE) across treatments, as the proportion of modelled cumulative net leaf C gain allocated to observed seedling C biomass.

@@

##Growth and morphology
Plant height, diameter and leaf area diverged between container volumes soon after start of the experiment (Figure `r allomfig`a-c). First to diverge was seedling leaf area, which significantly differed among soil volumes (*P* < 0.029) from the 5^th^ week of the experiment. Significant differences were then observed in both height (8^th^ week) and then diameter (9^th^ week) among soil volumes (*P* < 0.045 & 0.035, respectively). The large reductions in height gain and total leaf area in smaller compared to larger containers continued throughout the experiment. In this field study, colder temperatures and reductions in total PPFD per day (Figure `r airvarsfig`) likely slowed the growth of seedlings in the final weeks of the experiment. Seedlings maintained diameter growth throughout the experiment, although growth was marginal in smaller soil volumes in the final month. Final seedling height significantly increased with increasing soil volume (*P* < 0.001). Increases in both final stem diameter and cumulative leaf area (both *P* < 0.001) were found with increasing soil volume. We did not find a continuous effect of container size; rather, plant size was similar among smaller containers, and differences were driven mainly by the largest container and the Free seedling treatments (Figure `r allomfig`).

@@

\
Both *J*~max~ and *V*~cmax~ were significantly higher in Free seedlings (30 % and 26 %, respectively) than container-grown seedlings with little variation between container volume treatments (Table `r pvetab3`). Overall, the *J*~max~ / *V*~cmax~ ratio was conserved across all treatments (1.6&pm;0.02). Leaf dark respiration rates were not significantly different across soil volumes (Table `r pvetab3`). The *g*~1~ parameter, generated for each seedling from the Medlyn *et al.* [-@medlyn2011reconciling] optimal stomatal conductance model, was highest in the smaller containers and decreased in larger containers and the Free seedling treatment (Table `r pvetab3`).

@@

##Whole-plant C balance
Modelled cumulative net leaf C gain over 120 days varied three-fold across containers sizes (Table `r pvetab1`). The Free seedling control had five-fold greater modelled total C gain than the average of container treatments (213.6 vs 42.7&pm;8.2 g C). For seedlings in containers, an average of 67 &pm;0.01 % of modelled C gain was not allocated to observed biomass, compared to 59.2 % for the Free seedling control. Consequently, the estimated CUE of containerized seedlings was lower than the Free seedling control (Figure `r cuefig`). As a result, the observed reductions in leaf *A*~n~ with decreasing soil volume when integrated across the 120 day experiment were not large enough to explain the reduction in observed seedling biomass across the container size treatments.

@@

\
With high rates of *g*~s~, non-limiting leaf water potential ($\Psi$~pd~), and consistent leaf $\delta$^13^C across soil volume treatments, there was little evidence that water stress caused the reduction in *A*~sat~. This finding is consistent with other container size studies without drought treatments.  For example, reduced *A*~max~ in cotton seedlings grown at elevated CO~2~ was attributed to sink-limited feedback inhibition from inadequate rooting volume, not decreased *g*~s~ [@thomas1991root]. Similarly, severe reductions in leaf photosynthesis in coffee plants were not attributed to impacts of container size on leaf water potentials or *g*~s~ [@ronchi2006growth]. It is thus likely that reductions in *A*~sat~ of well-watered seedlings observed in our study of *E. tereticornis* seedlings are not the result of limiting soil moisture availability.

@@

\
Focusing only on empirical results in this study points to *A*~n~ driving reductions in seedling growth. Modelling cumulative net seedling C gain, however, suggests that other components of the C balance, beyond *A*~n~, are required to explain the observed seedling biomass response to sink manipulation. This finding is noteworthy, as partial accounting of the different mechanisms involved in plant C balance can lead to erroneous conclusions [@valentini2000respiration]. For example, the fraction of photosynthate used in respiration is known to vary depending on species and local environmental conditions [@Lambers2008], yet is often considered a static parameter in process based growth models. Thus, we agree with conclusions from Valentine and M&#228;kel&#228; [-@valentine2005bridging] that the problem with predicting tree growth is a problem in forecasting the assimilation and allocation of C and other constituents. Future empirical and modelling studies should focus on how feedbacks from sink activity affect both rates of *A*~n~ and the fate of C allocated at the whole plant level to growth, respiration, exudation and C storage. The degree to which these mechanisms regulate the C available for growth will undoubtedly shift across different species and environmental conditions.
